Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-10 newpaper/paper.tex
wip

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,(\msf{ask}\,\msf{MonadReaderDict})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(\vx + 20))
\end{alignat*}
Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is the
projection of the binding function. Here, a runtime closure will be created for
the $\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$
will create additional dynamic closures.

The difference between optimized and unoptimized code is already large here, and
it gets even larger when we consider monad transformers or code that is
polymorphic over monads. In Haskell, such code is pervasive, even in fairly
basic programs which do not use fancy abstractions. Consider the $\msf{mapM}$
function from the Haskell Prelude:
\begin{alignat*}{3}
  & \msf{mapM} :: \Monad\,\vm \Rightarrow (\va \to \vm\,\vb) \to [\va] \to \vm\,[\vb]
\end{alignat*}
This is a third-order and second-rank polymorphic function in disguise, because
its monad dictionary argument contains the polymorphic second-order method
$(\bind)$.  Compiling $\msf{mapM}$ efficiently relies on inlining the instance
dictionary, then inlining the methods contained there, and also inlining the
functions that the higher-order binding is applied to.

@@

In this paper we use staged compilation to address issues of robustness. The
idea is to shift as much as possible work from general-purpose optimization to
metaprograms.

@@

\begin{itemize}
\item In Section \ref{sec:overview-of-cftt} we present the two-level type theory
  CFTT, where the object level is first-order simply-typed and the meta level is
  dependently typed. The object language supports an operational semantics
  without runtime closures, and can be compiled with only statically known
  function calls. We provide a supplementary Agda formalization of the
  operational semantics of the object language.
\item In Section \ref{sec:monad-transformers} we build a monad transformer library. We believe
  that this is a good demonstration, because monads and monad transformers are
  the most widely used effect system in Haskell, and at the same time their
  compilation to efficient code can be surprisingly difficult. The main idea is
  to have monads and monad transformers \emph{only} in the metalanguage, and try
  to express as much as possible at the meta level, and funnel the object-meta
  interactions through specific \emph{binding-time improvements}.  We show how
  almost all definitions from the Haskell ecosystem of monad transformers can be
  reused in the meta level of CFTT. We also develop let-insertion, join points
  and case switching on object-level data.
\item In Section \ref{sec:stream-fusion} we build a pull-based stream fusion library. Here,
  we demonstrate essential usage of dependent types, in providing guaranteed
  fusion for arbitrary combinations of $\concatMap$ and $\mit{zip}$. We use a
  state machine representation that is based on \emph{sums-of-products} of
  object-level values. We show that CFTT is compatible with a
  \emph{generativity} axiom, which internalizes the fact that metaprograms
  cannot inspect the structure of object-level terms. We use this to show that
  the universe of sums-of-products is closed under $\Sigma$-types. This in turn
  enables a very concise definition of $\concatMap$.
\item
  We adapt the contents of the paper to typed Template Haskell
  \cite{DBLP:journals/pacmpl/XiePLWYW22}, with some modifications,
  simplifications and fewer guarantees about generated code.  In particular,
  Haskell does not have enough dependent types for the simple $\concatMap$
  definition, but we can still work around this limitation. We also provide a
  precise Agda embedding of CFTT and our libraries as described in this
  paper. Here, the object theory is embedded as a collection of postulated
  operations, and we can use Agda's normalization command to print out generated
  object code.
\end{itemize}

@@

$\bs{\Ty}$ is the universe of types in the object language. It is itself a
metatype, so so we have $\Ty : \MTy$. All construction and elimination rules of
type formers in $\Ty$ stay within $\Ty$. We further split $\Ty$ to two
sub-universes.

@@

With what we have seen so far, there is no interaction between the meta and
object levels. We make such interaction possible with \emph{staging primitives}.

@@

A CFTT program is a mixture of object-level and meta-level top-level definitions
and declarations. We use \emph{unstaging} to refer to the process of evaluating
metaprograms in a CFTT program in order to extract purely object-level code.
The output is guaranteed to be well-typed.

@@

@@

We give an overview of the unstaging algorithm together with its correctness.
We reuse the concepts and most of the results from \cite{staged2ltt}. Our main
task is to extend ibid.\,with identity types and inductive types in the
metalanguage, since we use them in this paper. For inductive types, it is the
most sensible to add W-types, since all inductive families can be faithfully
derived from them in our setting \cite{whynotw}.

This however requires a substantial amount of technical background from
\cite{staged2ltt}, so we relegate it to Appendix \ref{appendix:A}. Here we only
review the basics.
\begin{itemize}
\item The input of unstaging is a CFTT term with an object-level type, which
      only depends on object-level free variables. The output is a term in the
      object theory. In the other direction, there is an evident embedding of
      object-theoretic terms as CFTT terms.
\item \emph{Soundness} means that unstaging followed by embedding is the
      identity map, up to conversion. In other words, unstaging respects conversion
      of CFTT terms.
\item \emph{Stability} means that embedding followed by unstaging is the
      identity map. In other words, unstaging has no action on terms which are
      already purely object-level (i.e.\ contain no splices).
\end{itemize}
In \cite{staged2ltt}, there is one more property, called \emph{strictness},
which means that unstaging does not perform $\beta\eta$-conversion in the
object theory, but this holds trivially in our case since we do not have such
conversion rules.

@@

Axioms clearly block computation, so they are incompatible with the mentioned
notion of soundness. But we are only interested in equations which hold
definitionally during unstaging. Hence, we can do the following: first, we erase
all identity proofs and their transports from CFTT programs, then we proceed
with unstaging as in \cite{staged2ltt}. This erasure can be formalzied as a
syntactic translation \cite{next700} from CFTT to CFTT extended with the
equality reflection principle \cite{TODO}. Thus, soundness of staging for CFTT
holds up to erasure of identity proofs.

@@

In this section we build a library for monads and monad transformers. We believe
that this is a good demonstration of CFTT-s abilities, since monads are
ubiquitous in Haskell programming and they also introduce a great amount of
abstraction that should be optimized away.

@@

We start with some preparatory work before getting to monads. We saw that $\Up
\vA \to \Up \vB$ is usually preferable to $\Up(\vA \to \vB)$. The two types are
equivalent during unstaging, up to the runtime equivalence of object programs,
and we can convert back and forth in CFTT:
\begin{alignat*}{3}
  &\mup : \Up (\vA \to \vB) \to \Up \vA \to \Up \vB && \ind\ind \mdown : (\Up \vA \to \Up \vB) \to \Up (\vA \to \vB) \\
  &\mup\,\vf\,\va = \ql \spl \vf\, \spl \va\qr   && \ind\ind \mdown\,\vf = \ql \lam \va.\,\spl(\vf\,\ql \va \qr) \qr
\end{alignat*}
We cannot show internally, using propositional equality, that these functions are
inverses, since we do not have $\beta\eta$-rules for object functions; but we
will not need this proof in the rest of the paper.

In the staged compilation and partial evaluation literature, the term
\emph{binding time improvement} is used to refer to such conversions, where the
``improved'' version supports more compile-time computation
\cite{partial-evaluation}. A general strategy for generating efficient ``fused''
programs, is to try to work as much as possible with improved representations,
and only convert back to object code at points where runtime dependencies are
unavoidable. Let us look at binding-time-improvement for product types now:
\begin{alignat*}{3}
  &\mup : \Up (\vA,\,\vB) \to (\Up \vA,\,\Up \vB) && \ind\ind \mdown : (\Up \vA,\,\Up \vB) \to \Up(\vA,\,\vB) \\
  &\mup\,\vx = (\qt{\fst\,\spl \vx},\, \qt{\snd\,\spl \vx})   && \ind\ind \mdown\,(\vx,\,\vy) = \qt{(\spl \vx,\,\spl \vy)}
\end{alignat*}

Here we overload Haskell-style product type notation, both for types and the
pair constructor, at both levels. There is a problem with this conversion
though: $\mup$ uses $\vx : \Up(\vA,\,\vB)$ twice, which can increase code size
and duplicate runtime computations. For example, $\mdown\,({\mup\,\ql \vf\,\vx
  \qr})$ is staged to $\ql (\fst\,(\vf\,\vx),\,\snd\,(\vf\,\vx)) \qr$. It would
be safer to first let-bind an expression with type $\Up(\vA,\,\vB)$, and then
only use projections of the newly bound variable. This is called
\emph{let-insertion} in staged compilation. But it is impossible to use
let-insertion in $\mup$ because the return type is in $\MTy$, and we cannot
introduce object binders in meta-level code.  Fortunately, there is a principled
solution.

@@

The solution is to use a monad which extends $\MTy$ with the ability to
freely generate object-level code and introduce object binders.
\begin{alignat*}{3}
  & \newtype \Gen\,(\vA : \MTy) = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
\end{alignat*}
This is a monad in $\MTy$ in a standard sense:
\begin{alignat*}{3}
  &\mathrlap{\instance \Monad\,\Gen\,\where}\\
  &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,\unGen\,(\vf\,\va)\,\vk))
\end{alignat*}
From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
CFTT. We will use type classes in an informal way, without precisely specifying
how they work. However, type classes are used in essentially the same way in the
Agda and Haskell implementations, with modest technical differences.

From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be
also derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.
An inhabitant of $\Gen\,\vA$ can be viewed as an action whose effect is to produce
some object code. $\Gen$ can be only ``run'' when the result type is an object
type:
\begin{alignat*}{3}
  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
  &\runGen\,\vma = \unGen\,\vma\,\id
\end{alignat*}
We can let-bind object expressions in $\Gen$:
\begin{alignat*}{3}
  & \gen : \{\vA : \Ty\} \to \Up \vA \to \Gen\,(\Up \vA) \\
  & \gen\,\va = \Gen \dlr \lam \vk.\,\ql \letdef \vx : \vA := \spl \va; \spl(\vk\,\ql \vx \qr) \qr
\end{alignat*}
And also recursive definitions of computations:
\begin{alignat*}{3}
  & \genRec : \{\vA : \CTy\} \to (\Up \vA \to \Up \vA) \to \Gen\,(\Up \vA) \\
  & \genRec\,\vf = \Gen \dlr \lam \vk.\,\qt{\letrec \vx : \vA := \spl(\vf\,\qt{\vx}); \spl(\vk\,\qt{\vx})}
\end{alignat*}
Now, using do-notation, we may write $\mdo \{\vx \fro \gen\,\qt{10 + 10}; \vy \fro
\gen\,\qt{20 + 20}\};\return \qt{\vx + \vy}\}$, for a $\Gen\,(\Up \Nat)$
action. Running this with $\runGen$ yields $\qt{\letdef \vx := 10 + 10; \letdef \vy
  := 20 + 20; \vx + \vy}$. We can also define a ``safer'' binding-time improvement for
products, using let-insertion:
\begin{alignat*}{6}
  &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)            \hspace{5em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,\vx = \mdo &&(\va,\,\vb) \fro \vx\\
  &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})     &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
\end{alignat*}

Working in $\Gen$ is convenient, since we can freely generate object code and
also have access to the full metalanguage. Also, the point of staging is that
\emph{eventually} all metaprograms will be used for the purpose of code
generation, so we eventually $\runGen$ all of our actions. So why not
just always work in $\Gen$? The implicit nature of $\Gen$ may make it harder to
reason about the size and content of generated code. This is a bit similar to
the $\msf{IO}$ monad in Haskell, where eventually everything needs to run in
$\msf{IO}$, but we may not want to write all of our code in $\msf{IO}$.

@@

Let us start with the $\Maybe$ monad. We have $\data \Maybe\,\vA :=
\Nothing\,|\,\Just\,\vA$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
we do not have enough type formers in $\VTy$. We could try to use the following
type for binding:
\[ (\bind) : \Up(\Maybe\,\vA) \to (\Up\vA \to \Up(\Maybe\,\vB)) \to \Up(\Maybe\,\vB) \]
This works, but the definition necessarily uses runtime case splits on $\Maybe$
values, many of which could be optimized away during staging. Also, not having
a ``real'' monad is inconvenient for the purpose of code reuse.

@@

\subsubsection{Case splitting in monads}
We often want to case-split on object-level data inside a monadic action, and
perform different actions in different branches. At first, this may seem
problematic, because we cannot directly compute metaprograms from object-level
case splits. Fortunately, with a little bit more work, this is actually possible
in any $\MonadGen$, for any value type.

@@

\emph{Completeness} of unstaging is shown by straightforward induction on the
syntax of the object theory, the same way as in \cite{staged2ltt}, and we omit
it here.

>>>>

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
\end{alignat*}
Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
from the dictionary. We get a runtime closure from the $\lam \vb.\,...$
function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also produce additional
closures.

The difference between optimized and unoptimized code is already large here, and
it gets even larger when we consider monad transformers or code that is
polymorphic over monads. In Haskell, such code is pervasive, even in fairly
basic programs which do not use fancy abstractions. Consider the $\msf{mapM}$
function from the Haskell Prelude:
\begin{alignat*}{3}
  & \msf{mapM} :: \Monad\,\vm \Rightarrow (\va \to \vm\,\vb) \to [\va] \to \vm\,[\vb]
\end{alignat*}
This is a third-order rank-2 polymorphic function in disguise, because
its monad dictionary argument contains the polymorphic second-order method
$(\bind)$.  Compiling $\msf{mapM}$ efficiently relies on inlining the instance
dictionary, then inlining the methods contained there, and also inlining the
functions that the higher-order binding is applied to.

@@

In this paper we use staged compilation to address issues of robustness. The
idea is to shift as much work as possible from general-purpose optimization to
metaprograms.

@@

\begin{itemize}
\item In Section \ref{sec:overview-of-cftt} we present the two-level type theory
  CFTT, where the object level is first-order simply-typed and the meta level is
  dependently typed. The object language supports an operational semantics
  without runtime closures, and can be compiled with only statically known
  function calls. We provide a supplementary Agda formalization of the
  operational semantics of the object language.
\item
  In Section \ref{sec:monad-transformers} we build a staged library for monad
  transformers \cite{DBLP:conf/popl/LiangHJ95}. We believe that this is a good
  demonstration, because monads and monad transformers are the most widely used
  effect system in Haskell, and at the same time their compilation to efficient
  code can be surprisingly difficult. The continuation monad is well-known in
  staged compilation \cite{DBLP:conf/lfp/Bondorf92}, and staged state monads
  have also been used
  \cite{DBLP:conf/pepm/SwadiTKP06,DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11}.
  These works used specific monads as tools for domain-specific code
  generation. In contrast, we propose ubiquitous staging for general-purpose
  monadic programming, where users can write code that looks similar to monadic
  code in Haskell but with deterministic and robust compilation to efficient
  code.
\item In Section \ref{sec:stream-fusion} we build a pull-based stream fusion library. Here,
  we demonstrate essential usage of dependent types, in providing guaranteed
  fusion for arbitrary combinations of $\concatMap$ and $\mit{zip}$. We use a
  state machine representation that is based on \emph{sums-of-products} of
  object-level values. We show that CFTT is compatible with a
  \emph{generativity} axiom, which internalizes the fact that metaprograms
  cannot inspect the structure of object-level terms. We use this to show that
  the universe of sums-of-products is closed under $\Sigma$-types. This in turn
  enables a very concise definition of $\concatMap$.
\item
  We adapt the contents of the paper to typed Template Haskell
  \cite{DBLP:journals/pacmpl/XiePLWYW22}, with some modifications,
  simplifications and fewer guarantees about generated code.  In particular,
  Haskell does not have enough dependent types for the simple $\concatMap$
  definition, but we can still work around this limitation. We also provide a
  precise Agda embedding of CFTT and our libraries as described in this
  paper. Here, the object theory is embedded as a collection of postulated
  operations, and we can use Agda's normalization command to print out generated
  object code.
\end{itemize}

@@

$\bs{\Ty}$ is the universe of types in the object language. It is itself a
metatype, so we have $\Ty : \MTy$. All construction and elimination rules of
type formers in $\Ty$ stay within $\Ty$. We further split $\Ty$ to two
sub-universes.

@@

With what we have seen so far, there is no interaction between the meta and
object levels. We make such interaction possible with the following primitives.

@@

We use \textbf{unstaging} to refer to the process of extracting object code from
CFTT programs, by evaluating all metaprograms in splices. This term has been
sporadically used in the literature, e.g.\ by Odersky and Rompf
\cite{DBLP:journals/cacm/RompfO12} and also by Choi et
al.\ \cite{DBLP:conf/popl/ChoiAYT11} with a somewhat different meaning in a
multi-stage context. The main precursor to this paper used ``staging'' instead
of our ``unstaging'' \cite{staged2ltt}; we use the latter because the former
conflicts with other common usages of ``staging''.

@@

\subsubsection{Comparison to defunctionalization}
Defunctionalization is a program translation which represents higher-order
functions using only first-order functions and inductive data
\cite{DBLP:journals/lisp/Reynolds98a,DBLP:conf/ppdp/DanvyN01}. The idea is to
represent each closure as a constructor of an inductive type, and implement
closure application as a top-level first-order function which switches on all
possible closure constructors. Hence, like our unstaging algorithm,
defunctionalization produces first-order code from higher-order code. We summarize
the differences.
\begin{itemize}
\item Defunctionalization mainly supports \emph{compiler optimizations}, by making
  closures transparent to analysis. It is not an optimization by itself; it does
  not change the amount of dynamic control flow or dynamic allocations in a
  program. A dynamic closure call becomes a dynamic case-switch and a
  heap-allocated ``functional'' closure becomes a heap-allocated inductive
  constructor.
\item Staging mainly supports optimization by \emph{programmers} by providing
  control over code generation. Unstaging does not translate higher-order
  functions to anything in particular. Instead, it is the execution of
  higher-order metaprograms that produce first-order programs.
\end{itemize}

@@

We give an overview of the unstaging algorithm together with its correctness.
Our main task is to extend \cite{staged2ltt} with identity types and inductive
types in the metalanguage, since we use them in this paper. For inductive types,
it is the most sensible to add W-types, since all inductive families can be
faithfully derived from them in our setting \cite{whynotw}.

This however requires a substantial amount of technical background from
\cite{staged2ltt}, so we relegate it to Appendix \ref{appendix:A}. Here we only
give an overview.
\begin{itemize}
\item The input of unstaging is a CFTT term with an object-level type, which
      only depends on object-level free variables. The output is a term in the
      object theory. In the other direction, there is an evident embedding of
      object-theoretic terms as CFTT terms.
\item \emph{Soundness} means that unstaging followed by embedding is the
      identity map, up to conversion. In other words, unstaging respects conversion
      of CFTT terms.
\item \emph{Stability} means that embedding followed by unstaging is the
      identity map. In other words, unstaging has no action on terms which are
      already purely object-level (i.e.\ contain no splices).
\end{itemize}
In \cite{staged2ltt}, there is one more property, called \emph{strictness},
which means that unstaging does not perform $\beta\eta$-conversion in the
object theory, but this holds trivially in our case since we do not have such
conversion rules.

@@

Axioms clearly block computation, so they are incompatible with the mentioned
notion of soundness. But we are only interested in equations which hold
definitionally during unstaging. Hence, we can do the following: first, we erase
all identity proofs and their transports from CFTT programs, then we proceed
with unstaging as in \cite{staged2ltt}. This erasure can be formalzied as a
syntactic translation \cite{next700} from CFTT to CFTT extended with the
equality reflection principle, which says that propositional equality implies
definitional equality (see e.g.\ \cite{hofmann95conservativity}). Thus,
soundness of staging for CFTT holds up to erasure of identity proofs.

@@

In this section we build a library for monads and monad transformers. We believe
that this is a good demonstration of CFTT's abilities, since monads are
ubiquitous in Haskell programming and they also introduce a great amount of
abstraction that should be optimized away.

@@

We do a preliminary overview before getting to monads. In the staged compilation
and partial evaluation literature, the term \emph{binding time improvement} is
used to refer to such conversions, where the ``improved'' version supports more
compile-time computation \cite[Chapter 12]{partial-evaluation}.

Translation from $\Up \vA \to \Up \vB$ to $\Up(\vA \to \vB)$ is a basic
binding-time improvement, as an $\eta$-expansion \cite{eta-expansion-trick}. The
two types are equivalent during unstaging, up to the runtime equivalence of
object programs, and we can convert back and forth in CFTT, the same way as in
MetaML \cite[Section 9]{metaml}:
\begin{alignat*}{3}
  &\mup : \Up (\vA \to \vB) \to \Up \vA \to \Up \vB && \ind\ind \mdown : (\Up \vA \to \Up \vB) \to \Up (\vA \to \vB) \\
  &\mup\,\vf\,\va = \ql \spl \vf\, \spl \va\qr   && \ind\ind \mdown\,\vf = \ql \lam \va.\,\spl(\vf\,\ql \va \qr) \qr
\end{alignat*}
We cannot show internally, using propositional equality, that these functions are
inverses, since we do not have $\beta\eta$-rules for object functions; but we
will not need this proof in the rest of the paper.

 A general strategy for generating efficient ``fused''
programs, is to try to work as much as possible with improved representations,
and only convert back to object code at points where runtime dependencies are
unavoidable. Let us look at binding-time-improvement for product types now:
\begin{alignat*}{3}
  &\mup : \Up (\vA,\,\vB) \to (\Up \vA,\,\Up \vB) && \ind\ind \mdown : (\Up \vA,\,\Up \vB) \to \Up(\vA,\,\vB) \\
  &\mup\,\vx = (\qt{\fst\,\spl \vx},\, \qt{\snd\,\spl \vx})   && \ind\ind \mdown\,(\vx,\,\vy) = \qt{(\spl \vx,\,\spl \vy)}
\end{alignat*}

Here we overload Haskell-style product type notation, both for types and the
pair constructor, at both levels. There is a problem with this conversion
though: $\mup$ uses $\vx : \Up(\vA,\,\vB)$ twice, which can increase code size
and duplicate runtime computations. For example, $\mdown\,({\mup\,\ql \vf\,\vx
  \qr})$ is staged to $\ql (\fst\,(\vf\,\vx),\,\snd\,(\vf\,\vx)) \qr$. It would
be safer to first let-bind an expression with type $\Up(\vA,\,\vB)$, and then
only use projections of the newly bound variable. This is called
\emph{let-insertion} in staged compilation. But it is impossible to use
let-insertion in $\mup$ because the return type is in $\MTy$, and we cannot
introduce object binders in meta-level code.

@@

A principled solution to the previous issue is to write code generators in
continuation-passing style, as first proposed by Bondorf
\cite{DBLP:conf/lfp/Bondorf92}. This can be structured as a continuation monad,
and later work used explicit monadic notation for it
\cite{DBLP:conf/emsoft/KiselyovST04,DBLP:journals/scp/CaretteK11,DBLP:conf/pepm/SwadiTKP06}. Our
definition is as follows:
\begin{alignat*}{3}
  & \newtype \Gen\,(\vA : \MTy) = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
\end{alignat*}
This is a monad in $\MTy$ in a standard sense:
\begin{alignat*}{3}
  &\mathrlap{\instance \Monad\,\Gen\,\where}\\
  &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,\unGen\,(\vf\,\va)\,\vk))
\end{alignat*}
In \cite{DBLP:conf/emsoft/KiselyovST04},\cite{DBLP:journals/scp/CaretteK11} and
\cite{DBLP:conf/pepm/SwadiTKP06}, the answer type is a parameter to the
continuation monad, while we have it as polymorphic. We have found that
polymorphic answer types are more convenient, because we do not have to
anticipate the type of the code output when are defining a code generator.
Hence, we can ``run'' actions as follows:
\begin{alignat*}{3}
  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
  &\runGen\,\vma = \unGen\,\vma\,\id
\end{alignat*}
From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
CFTT. We will use type classes in an informal way, without precisely specifying
how they work. However, type classes are used in essentially the same way in the
Agda and Haskell implementations, with modest technical differences.  From the
$\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be also
derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.

We can let-bind object expressions in $\Gen$:
\begin{alignat*}{3}
  & \gen : \{\vA : \Ty\} \to \Up \vA \to \Gen\,(\Up \vA) \\
  & \gen\,\va = \Gen \dlr \lam \vk.\,\ql \letdef \vx : \vA := \spl \va; \spl(\vk\,\ql \vx \qr) \qr
\end{alignat*}
And also recursive definitions of computations:
\begin{alignat*}{3}
  & \genRec : \{\vA : \CTy\} \to (\Up \vA \to \Up \vA) \to \Gen\,(\Up \vA) \\
  & \genRec\,\vf = \Gen \dlr \lam \vk.\,\qt{\letrec \vx : \vA := \spl(\vf\,\qt{\vx}); \spl(\vk\,\qt{\vx})}
\end{alignat*}
Now, using $\mbf{do}$-notation, we may write $\mdo \{\vx \fro \gen\,\qt{10 + 10}; \vy \fro
\gen\,\qt{20 + 20}\};\return \qt{\vx + \vy}\}$, for a $\Gen\,(\Up \Nat)$
action. Running this with $\runGen$ yields $\qt{\letdef \vx := 10 + 10; \letdef \vy
  := 20 + 20; \vx + \vy}$. We can also define a ``safer'' binding-time improvement for
products, using let-insertion:
\begin{alignat*}{6}
  &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)            \hspace{5em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,\vx = \mdo &&(\va,\,\vb) \fro \vx\\
  &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})     &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
\end{alignat*}

Working in $\Gen$ is convenient, since we can freely generate object code and
also have access to the full metalanguage. Also, the point of staging is that
\emph{eventually} all metaprograms will be used for the purpose of code
generation, so we eventually $\runGen$ all of our actions. So why not just
always work in $\Gen$? The implicit nature of $\Gen$ may make it harder to
reason about the size and content of generated code. This is a bit similar to
the $\msf{IO}$ monad in Haskell \cite{jones2001tackling}, where eventually
everything needs to run in $\msf{IO}$, but we may not want to write all of our
code in $\msf{IO}$.

@@

Let us start with the $\Maybe$ monad. We have $\data \Maybe\,\vA :=
\Nothing\,|\,\Just\,\vA$, and $\Maybe$ itself is available as a $\VTy \to \VTy$
metafunction. However, we cannot directly fashion a monad out of $\Maybe$, since
we do not have enough type formers in $\VTy$. In particular, we cannot store
functions inside $\Maybe$ which rules out the applicative $\ap$ operation.  We
could try to use the following type for binding:
\[ (\bind) : \Up(\Maybe\,\vA) \to (\Up\vA \to \Up(\Maybe\,\vB)) \to \Up(\Maybe\,\vB) \]
This works, but the definition necessarily uses runtime case splits on $\Maybe$
values, many of which could be optimized away during staging. Also, not having
a ``real'' monad is inconvenient for the purpose of code reuse.

@@

\subsubsection{Case splitting in monads}
We often want to case-split on object-level data inside a monadic action, and
perform different actions in different branches. At first, this may seem
problematic, because we cannot directly compute metaprograms from object-level
case splits. Fortunately, with a little bit more work, this is possible in any
$\MonadGen$, for any value type, using a variation of the so-called ``trick''
\cite{eta-expansion-trick}.

@@

\emph{Strictness} of unstaging holds trivially, since there are no $\beta$ and
$\eta$ rules in the object theory. \emph{Completeness} of unstaging is shown by
straightforward induction on the syntax of the object theory, the same way as in
\cite{staged2ltt}, and we omit it here.
